Mental Health Factors and Covid

### 2025-05-01 code/analysis-code/analysis.qmd
Part 6 updates

To start, I'll load packages. As I need packages along the way, I'll return here and add them.
```{r}
library(tidyverse)
library(survey)
library(tableone)
library(mice)
library(mitools)
library(miceadds)
library(tidymodels)
library(caret)
library(gt)
library(pROC)
library(ranger)
library(glmnet)
library(Metrics)
library(purrr)
library(here)
```

@@

Next, we will make the multivariate model. I will use cross validation for this. The pooled AUC across imputations is found to be 0.5230.
```{r}
# Define the AUC cross-validation function for imputed datasets
cross_validation_imputed_auc <- function(imputed_data, predictors, k_folds = 5) {
  auc_list <- numeric(length(imputed_data))  # Store AUC for each imputation

  for (i in 1:length(imputed_data)) {
    data <- imputed_data[[i]]

    # drop rows with missing values in COVID, predictors, or SAMPWEIGHT
    data <- data %>% drop_na(all_of(c("COVID", predictors, "SAMPWEIGHT")))

@@

    fold_auc <- numeric(k_folds)  # AUC for each fold

@@

      # Fit random forest model on training data
      formula <- as.formula(paste("COVID ~", paste(predictors, collapse = " + ")))
      rf_model <- ranger(
        formula = formula,
        data = train_data,
        case.weights = train_data$SAMPWEIGHT,
        probability = TRUE
      )

      # Predict probabilities for class 1 (COVID = 1)
      pred_probs <- predict(rf_model, data = test_data)$predictions[, 2]
      actual <- test_data$COVID

@@

    # Average AUC over folds for this imputed dataset
    auc_list[i] <- mean(fold_auc)
  }

  # Average AUC over all imputed datasets
  overall_auc <- mean(auc_list)

@@

# Assuming you have your imputed_data and a vector of predictors
results <- cross_validation_imputed_auc(imputed_data = data_imputed,
                                        predictors = c("EXERCISE", "SLEEP", "DEPRESSED", "TROUBLE_SLEEPING", "SOCIAL"))

# View results
results$auc_per_imputation  # AUC for each imputed dataset
results$overall_auc         # Average AUC across imputations
```

Now, we will make a LASSO regression model. For this, I'm going to use the multiple imputation dataset with cross validation with 5 folds. Instead of using complex survey design, I will weight the model by sampleweight. The pooled AUC is  0.529.

```{r}
# Extract the first completed imputed dataset
data1 <- data_imputed[[1]]

# Define the survey design
survey_design <- svydesign(
  ids = ~PSU,
  strata = ~STRATA,
  weights = ~SAMPWEIGHT,
  data = data1,
  nest=TRUE
)

set.seed(1234)  # for reproducibility

# Choose outcome and predictors
outcome_var <- "COVID"
predictor_vars <- c("EXERCISE", "SLEEP", "DEPRESSED", "TROUBLE_SLEEPING", "SOCIAL")

# Loop over imputations and compute AUC
auc_list <- map(data_imputed, function(dat) {
  # Drop missing in predictors or outcome if any slipped through
  dat <- dat %>% drop_na(all_of(c(outcome_var, predictor_vars, "SAMPWEIGHT")))

  # Create design matrix X and response y
  X <- model.matrix(as.formula(paste(outcome_var, "~", paste(predictor_vars, collapse = "+"))), dat)[, -1]
  y <- dat[[outcome_var]]
  w <- dat$SAMPWEIGHT

  # Split into training and test sets (e.g., 70/30 split)
  train_idx <- sample(seq_len(nrow(X)), size = 0.7 * nrow(X))
  test_idx <- setdiff(seq_len(nrow(X)), train_idx)

@@

  X_test <- X[test_idx, ]
  y_test <- y[test_idx]
  w_test <- w[test_idx]

  # Fit random forest model (instead of lasso, because it's a classification problem)
  rf_model <- ranger(
    formula = as.formula(paste(outcome_var, "~", paste(predictor_vars, collapse = "+"))),
    data = dat[train_idx, ],
    case.weights = w_train,
    probability = TRUE
  )

  # Predict probabilities for class 1
  pred_probs <- predict(rf_model, data = dat[test_idx, ])$predictions[,2]

  # Compute AUC
  auc_result <- roc(y_test, pred_probs)$auc
  return(auc_result)
})

# Pool AUC across imputations
mean_auc <- mean(unlist(auc_list))

# Output
cat("AUC for each imputation:\n")
print(unlist(auc_list))
cat("\nPooled AUC across imputations:", round(mean_auc, 3), "\n")
```

Finally, we will construct a random forest model. For this, I will not use the multiple imputation dataset, as before, I will use the first imputed dataset, with cross validation 5 folds. The pooled AUC is 0.556.

```{r}
library(ranger)
library(caret)
library(pROC)
library(dplyr)

set.seed(1234)
k <- 5
auc_list <- numeric(length(data_imputed))

for (i in seq_along(data_imputed)) {
  data1 <- data_imputed[[i]]

  # Remove missing outcome
  data1_complete <- data1 %>% filter(!is.na(COVID))

  # Make sure outcome is binary (0/1)
  data1_complete$COVID <- as.numeric(data1_complete$COVID)

  # Add weights column if not present
  data1_complete$weights <- data1_complete$SAMPWEIGHT

  auc_folds <- numeric(k)

    train_data <- data1_complete[train_idx, ]
    test_data  <- data1_complete[test_idx, ]

    # Fit random forest classifier
    rf_model <- ranger(
      formula = COVID ~ EXERCISE + SLEEP + DEPRESSED + TROUBLE_SLEEPING + SOCIAL, #formula
      data = train_data,
      case.weights = train_data$weights,
      probability = TRUE
    )

    # Predict probabilities for class 1
    pred_probs <- predict(rf_model, data = test_data)$predictions[,2]

    # Compute AUC
    auc_folds[fold_idx] <- roc(test_data$COVID, pred_probs, quiet = TRUE)$auc
  }

  # Store average AUC for this imputation
  auc_list[i] <- mean(auc_folds)
}

# Pool AUC across imputations
mean_auc <- mean(auc_list)
print(paste("Pooled AUC across imputations:", round(mean_auc, 3)))
```

@@

# Save the table as a PNG
here()
gtsave(model_table, here("results", "figures", "model_table.png"))
```

>>>>

To start, I'll load packages. As I need packages along the way, I'll return here and add them.
```{r}
library(tidyverse)
library(survey)
library(tableone)
library(mice)
library(mitools)
library(miceadds)
library(tidymodels)
library(caret)
library(gt)
library(scales)
library(pROC)
library(ranger)
library(glmnet)
library(Metrics)
library(purrr)
library(here)
```

@@

Next, we will make the multivariate model. I will use cross validation for this.Pooled AUC is calculated and a observed-expected plot is generated.
```{r}
library(pROC)
library(dplyr)

cross_validation_imputed_auc_logit <- function(imputed_data, predictors, k_folds = 5) {
  auc_list <- numeric(length(imputed_data))  # Store AUC for each imputation

  for (i in seq_along(imputed_data)) {
    data <- imputed_data[[i]]

    # Drop rows with missing values in outcome, predictors, or weights
    data <- data %>% drop_na(all_of(c("COVID", predictors, "SAMPWEIGHT")))

@@

    fold_auc <- numeric(k_folds)

@@

      # Fit logistic regression model
      formula <- as.formula(paste("COVID ~", paste(predictors, collapse = " + ")))
      logit_model <- glm(
        formula = formula,
        data = train_data,
        family = binomial(),
        weights = SAMPWEIGHT
      )

      # Predict probabilities on test data
      pred_probs <- predict(logit_model, newdata = test_data, type = "response")
      actual <- test_data$COVID

@@

    auc_list[i] <- mean(fold_auc)
  }

  overall_auc <- mean(auc_list)

@@

# Usage
results_logit <- cross_validation_imputed_auc_logit(
  imputed_data = data_imputed,
  predictors = c("EXERCISE", "SLEEP", "DEPRESSED", "TROUBLE_SLEEPING", "SOCIAL")
)

# View results
results_logit$auc_per_imputation
results_logit$overall_auc

# Function to compute calibration data per imputation
get_calibration_data <- function(data, predictors, n_bins = 10) {
  # Fit logistic regression model
  formula <- as.formula(paste("COVID ~", paste(predictors, collapse = " + ")))
  model <- glm(formula, data = data, family = binomial(), weights = SAMPWEIGHT)

  # Predict probabilities
  data$predicted <- predict(model, type = "response")

  # Bin predictions into deciles
  data <- data %>%
    mutate(bin = ntile(predicted, n_bins)) %>%
    group_by(bin) %>%
    summarise(
      mean_pred = mean(predicted),
      obs_rate = mean(COVID),
      .groups = "drop"
    )

  return(data)
}

# Combine calibration data across imputations
all_calib_data <- purrr::map_dfr(
  .x = seq_along(data_imputed),
  .f = function(i) {
    data <- data_imputed[[i]] %>% 
      drop_na(COVID, EXERCISE, SLEEP, DEPRESSED, TROUBLE_SLEEPING, SOCIAL, SAMPWEIGHT)

    calib <- get_calibration_data(data, predictors = c("EXERCISE", "SLEEP", "DEPRESSED", "TROUBLE_SLEEPING", "SOCIAL"))
    calib$imputation <- paste0("Imputation ", i)
    return(calib)
  }
)

# Plot calibration curves
LogReg_plot <- ggplot(all_calib_data, aes(x = mean_pred, y = obs_rate, color = imputation)) +
  geom_line() +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous("Expected Probability", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous("Observed Proportion", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(title = "Calibration Plot: Logistic Regression (All 5 Imputations)", color = "Imputation") +
  theme_minimal()

# Save the plot using the `here` package
ggsave(filename = here::here("results", "figures", "log_calibration_plot.png"),
       plot = LogReg_plot,
       width = 8, height = 6, dpi = 300)

Now, we will make a LASSO regression model. For this, I'm going to use the multiple imputation dataset with cross validation with 5 folds. Instead of using complex survey design, I will weight the model by sampleweight. Pooled AUC is calculated and a observed-expected plot is generated.

```{r}
library(glmnet)
library(caret)
library(Metrics)  # for rmse()
library(dplyr)

set.seed(1234)
k <- 5
auc_list <- c()

# Loop through 5 imputed datasets
for (i in 1:5) {
  data1 <- data_imputed[[i]]

  # Remove rows with missing outcome
  data1_complete <- data1 %>% filter(!is.na(COVID))

  preds <- rep(NA, nrow(data1_complete))

    train_data <- data1_complete[train_idx, ] %>% drop_na()
    test_data  <- data1_complete[test_idx, ] %>% drop_na()

    # Create model matrix (remove outcome column)
    x_train <- model.matrix(COVID ~ ., data = train_data)[, -1]
    y_train <- train_data$COVID

    x_test <- model.matrix(COVID ~ ., data = test_data)[, -1]

    # Fit Lasso with CV
    lasso_fit <- cv.glmnet(x = x_train, y = y_train, alpha = 1, family = "binomial")

    # Predict probabilities (type = "response" for prob)
    preds[test_idx] <- predict(lasso_fit, newx = x_test, s = "lambda.min", type = "response")
  }

  actuals <- data1_complete$COVID

  # AUC using pROC
  roc_obj <- roc(actuals, preds)
  fold_auc <- auc(roc_obj)
  auc_list[i] <- fold_auc
}

pooled_auc <- mean(auc_list)
print(pooled_auc)

################################################################################
## plot expected vs observed

set.seed(1234)
k <- 5
all_preds_df <- data.frame()

for (i in 1:5) {
  data1 <- data_imputed[[i]]
  data1_complete <- data1 %>% filter(!is.na(COVID))

  folds <- createFolds(data1_complete$COVID, k = k, list = TRUE, returnTrain = FALSE)

  preds <- rep(NA, nrow(data1_complete))

    train_data <- data1_complete[train_idx, ] %>% drop_na()
    test_data  <- data1_complete[test_idx, ] %>% drop_na()

    x_train <- model.matrix(COVID ~ ., data = train_data)[, -1]
    y_train <- train_data$COVID
    x_test  <- model.matrix(COVID ~ ., data = test_data)[, -1]

    lasso_fit <- cv.glmnet(x = x_train, y = y_train, alpha = 1, family = "binomial")

    preds[test_idx] <- predict(lasso_fit, newx = x_test, s = "lambda.min", type = "response")
  }

  all_preds_df <- bind_rows(all_preds_df, data.frame(
    imputation = i,
    predicted = preds,
    observed = data1_complete$COVID
  ))
}

# Bin predictions into 10 bins and compute calibration stats
calib_data <- all_preds_df %>%
  mutate(bin = ntile(predicted, 10)) %>%
  group_by(imputation, bin) %>%
  summarise(
    mean_pred = mean(predicted, na.rm = TRUE),
    obs_rate = mean(observed, na.rm = TRUE),
    .groups = "drop"
  )

Lasso_plot <- ggplot(calib_data, aes(x = mean_pred, y = obs_rate, color = factor(imputation))) +
  geom_line() +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous("Expected Probability", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous("Observed Proportion", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(
    title = "Calibration Plot: Lasso Logistic Regression (All 5 Imputations)",
    color = "Imputation"
  ) +
  theme_minimal()

print(Lasso_plot)

# Save the plot using the `here` package
ggsave(filename = here::here("results", "figures", "lasso_calibration_plot.png"),
       plot = Lasso_plot,
       width = 8, height = 6, dpi = 300)

################################################################################
# Keep variable names for predictors
formula <- as.formula("COVID ~ .")

# Explicitly remove rows with any NA in predictors (even if COVID is not NA)
train_data <- train_data %>% select(COVID, everything()) %>% drop_na()

# Now get model matrix
x_train <- model.matrix(formula, data = train_data)[, -1]
y_train <- train_data$COVID

nrow(x_train)
length(y_train)
```

```{r}
# Load required libraries
library(glmnet)      # For Lasso regression
library(pROC)        # For AUC calculation
library(ggplot2)     # For plotting
library(dplyr)       # For data manipulation
library(mice)        # For working with imputed data
library(tidyr)       # For data reshaping

# --- LASSO MODEL WITH CROSS-VALIDATION ON IMPUTED DATA ---

# Function to run Lasso with sample weights and calculate performance metrics
run_lasso_cv <- function(imp_data, outcome = "COVID", 
                        predictors = c("SLEEP", "DEPRESSED", "SOCIAL", "TROUBLE_SLEEPING", "EXERCISE"),
                        weight_var = "sampleweight") {

  # Extract data from current imputation
  X <- as.matrix(imp_data[, predictors])
  y <- imp_data[[outcome]]
  weights <- imp_data[[weight_var]]

  # Standardize predictors (important for Lasso)
  X_scaled <- scale(X)

  # Run cross-validated Lasso with sample weights
  set.seed(123) # For reproducibility
  cv_model <- cv.glmnet(
    x = X_scaled,
    y = y,
    family = "binomial",
    alpha = 1,       # 1 = Lasso
    weights = weights,
    nfolds = 5,
    type.measure = "auc"
  )

  # Get the best lambda (penalty parameter)
  best_lambda <- cv_model$lambda.min

  # Refit model with best lambda
  final_model <- glmnet(
    x = X_scaled,
    y = y,
    family = "binomial",
    alpha = 1,
    weights = weights,
    lambda = best_lambda
  )

  # Get predictions
  pred_probs <- predict(final_model, newx = X_scaled, type = "response")

  # Calculate AUC
  roc_obj <- roc(y, pred_probs)
  auc_value <- auc(roc_obj)

  # Create calibration data for expected vs observed plot
  n_groups <- 10  # Number of bins for calibration plot

  # Create calibration data
  calib_data <- data.frame(
    predicted = as.vector(pred_probs),
    observed = y
  )

  # Sort by predicted probability and divide into bins
  calib_data <- calib_data[order(calib_data$predicted), ]
  calib_data$bin <- cut(seq(1, nrow(calib_data)), breaks = n_groups, labels = FALSE)

  # Calculate mean predicted and observed values for each bin
  calibration <- calib_data %>%
    group_by(bin) %>%
    summarize(
      n = n(),
      mean_predicted = mean(predicted),
      mean_observed = mean(observed)
    )

  # Return results
  return(list(
    model = final_model,
    lambda = best_lambda,
    auc = auc_value,
    calibration = calibration,
    coefs = coef(final_model)
  ))
}

# ----------------------------
# MAIN ANALYSIS
# ----------------------------

# Check the structure of the imputed data
print(class(data_imputed))

# Initialize results containers
auc_values <- numeric(length = 5)
calibration_results <- list()
coef_results <- list()

# Function to run Lasso with sample weights and calculate performance metrics
run_lasso_cv <- function(imp_data, outcome = "COVID", 
                        predictors = c("SLEEP", "DEPRESSED", "SOCIAL", "TROUBLE_SLEEPING", "EXERCISE"),
                        weight_var = "sampleweight") {

  # Extract data from current imputation and handle missing values
  # First, check for and handle missing data
  complete_cases <- complete.cases(imp_data[, c(outcome, predictors, weight_var)])
  if (!all(complete_cases)) {
    message("Removing ", sum(!complete_cases), " rows with missing values")
    imp_data <- imp_data[complete_cases, ]
  }

  X <- as.matrix(imp_data[, predictors])
  y <- imp_data[[outcome]]
  weights <- imp_data[[weight_var]]

  # Standardize predictors (important for Lasso)
  X_scaled <- scale(X)

  # Run cross-validated Lasso with sample weights
  set.seed(123) # For reproducibility
  cv_model <- cv.glmnet(
    x = X_scaled,
    y = y,
    family = "binomial",
    alpha = 1,       # 1 = Lasso
    weights = weights,
    nfolds = 5,
    type.measure = "auc"
  )

  # Get the best lambda (penalty parameter)
  best_lambda <- cv_model$lambda.min

  # Refit model with best lambda
  final_model <- glmnet(
    x = X_scaled,
    y = y,
    family = "binomial",
    alpha = 1,
    weights = weights,
    lambda = best_lambda
  )

# Combine calibration results from all imputations
all_calibration <- do.call(rbind, calibration_results)

# ----------------------------
# OUTPUT RESULTS
# ----------------------------

# Print AUC values for each imputation
cat("\nAUC values for each imputation:\n")
print(auc_values)
cat("\nMean AUC across imputations:", mean(auc_values), "\n")

# Plot expected vs observed probabilities for all imputations
calibration_plot <- ggplot(all_calibration, aes(x = mean_predicted, y = mean_observed, color = factor(imputation))) +
  geom_point(size = 3) +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(
    title = "Expected vs. Observed Probabilities",
    subtitle = paste("Mean AUC =", round(mean(auc_values), 3)),
    x = "Expected (Predicted Probability)",
    y = "Observed (Actual Proportion)",
    color = "Imputation"
  ) +
  theme_minimal() +
  coord_equal() +
  xlim(0, 1) + 
  ylim(0, 1)

print(calibration_plot)

# Print average coefficients across imputations
cat("\nAverage Lasso coefficients across imputations:\n")
# Extract coefficient names from the first imputation
coef_names <- rownames(coef_results[[1]])

# Create matrix to store coefficients from all imputations
all_coefs <- matrix(0, nrow = length(coef_names), ncol = 5)
rownames(all_coefs) <- coef_names

# Fill in the coefficient matrix
for (i in 1:5) {
  all_coefs[, i] <- as.vector(coef_results[[i]])
}

# Calculate mean coefficients
mean_coefs <- rowMeans(all_coefs)
coef_summary <- data.frame(
  Variable = coef_names,
  Mean_Coefficient = mean_coefs
)

# Print non-zero coefficients
non_zero_coefs <- coef_summary[abs(coef_summary$Mean_Coefficient) > 0.001, ]
print(non_zero_coefs[order(abs(non_zero_coefs$Mean_Coefficient), decreasing = TRUE), ])
```

```{r}
library(dplyr)
library(ggplot2)
library(scales)

calib_data_list <- vector("list", length(data_imputed))

for (i in seq_along(data_imputed)) {
  dat <- data_imputed[[i]] %>%
    drop_na(all_of(c(outcome_var, predictor_vars, "SAMPWEIGHT")))

  X <- model.matrix(as.formula(paste(outcome_var, "~", paste(predictor_vars, collapse = "+"))), dat)[, -1]
  y <- dat[[outcome_var]]
  w <- dat$SAMPWEIGHT

  set.seed(1234 + i)
  train_idx <- sample(seq_len(nrow(X)), size = 0.7 * nrow(X))
  test_idx <- setdiff(seq_len(nrow(X)), train_idx)

@@

  X_test <- X[test_idx, ]
  y_test <- y[test_idx]

  cv_fit <- cv.glmnet(
    x = X_train,
    y = y_train,
    weights = w_train,
    alpha = 1,
    family = "binomial",
    standardize = TRUE
  )

  pred_probs <- predict(cv_fit, newx = X_test, s = "lambda.min", type = "response") %>% as.vector()

  # Count unique predicted probabilities
  n_unique_preds <- length(unique(pred_probs))

  # Only proceed with binning if there are at least 2 unique predicted probabilities
  if (n_unique_preds >= 2) {
    # Safely bin the predicted probabilities
    n_bins <- min(5, n_unique_preds)

    calib_data <- tibble(
      actual = y_test,
      predicted = pred_probs
    ) %>%
      mutate(bin = cut_number(predicted, n = n_bins, labels = FALSE)) %>%
      group_by(bin) %>%
      summarise(
        mean_pred = mean(predicted),
        obs_rate = mean(actual),
        .groups = "drop"
      ) %>%
      mutate(imputation = paste0("Imputation ", i))

    calib_data_list[[i]] <- calib_data
  } else {
    message("Skipping calibration plot for imputation ", i, ": too few unique predicted values (", n_unique_preds, " unique values)")
  }
}

# Combine all calibration data
all_calib_data <- bind_rows(calib_data_list)

# Create calibration plot
ggplot(all_calib_data, aes(x = mean_pred, y = obs_rate, color = imputation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous("Expected Probability", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous("Observed Proportion", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(title = "Calibration Plot: Lasso Logistic Regression", color = "Imputation") +
  theme_minimal()

# Combine calibration data
all_calib_data <- bind_rows(calib_data_list)

# Plot calibration curve
ggplot(all_calib_data, aes(x = mean_pred, y = obs_rate, color = imputation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous("Expected Probability", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous("Observed Proportion", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(title = "Calibration Plot: Lasso Logistic Regression", color = "Imputation") +
  theme_minimal()

# Combine calibration data
all_calib_data <- bind_rows(calib_data_list)

# Plot calibration curve
ggplot(all_calib_data, aes(x = mean_pred, y = obs_rate, color = imputation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous("Expected Probability", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous("Observed Proportion", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(title = "Calibration Plot: Lasso Logistic Regression", color = "Imputation") +
  theme_minimal()
# Save the plot
ggsave(filename = here::here("results", "figures", "lasso_calibration_plot.png"),
       plot = calibration_plot_lasso,
       width = 8, height = 6, dpi = 300)

# Initialize list to store AUCs
auc_list <- vector("numeric", length(data_imputed))

for (i in seq_along(data_imputed)) {
  dat <- data_imputed[[i]] %>%
    drop_na(all_of(c(outcome_var, predictor_vars, "SAMPWEIGHT")))

  X <- model.matrix(as.formula(paste(outcome_var, "~", paste(predictor_vars, collapse = "+"))), dat)[, -1]
  y <- dat[[outcome_var]]
  w <- dat$SAMPWEIGHT

  set.seed(1234 + i)
  train_idx <- sample(seq_len(nrow(X)), size = 0.7 * nrow(X))
  test_idx <- setdiff(seq_len(X), train_idx)

  X_test <- X[test_idx, ]
  y_test <- y[test_idx]

  # Fit lasso model
  cv_fit <- cv.glmnet(
    x = X_train,
    y = y_train,
    weights = w_train,
    alpha = 1,
    family = "binomial",
    standardize = TRUE
  )

  # Predict probabilities
  pred_probs <- predict(cv_fit, newx = X_test, s = "lambda.min", type = "response") %>% as.vector()

  # Compute AUC
  auc_val <- roc(y_test, pred_probs)$auc
  auc_list[i] <- auc_val
}

# Print AUCs
cat("AUC for each imputation:\n")
print(round(auc_list, 3))

# Mean (pooled) AUC
cat("\nPooled AUC across imputations:", round(mean(auc_list), 3), "\n")
```

Finally, we will construct a random forest model. For this, I will not use the multiple imputation dataset, as before, I will use the first imputed dataset, with cross validation 5 folds. Pooled AUC is calculated and a observed-expected plot is generated.

```{r}
cross_validation_imputed_predictions <- function(imputed_data, predictors, k_folds = 5) {
  auc_list <- numeric(length(imputed_data))  # Store AUCs
  predictions_list <- vector("list", length(imputed_data))  # Store predictions and actuals

  for (i in 1:length(imputed_data)) {
    data <- imputed_data[[i]] %>%
      drop_na(all_of(c("COVID", predictors, "SAMPWEIGHT")))

    fold_auc <- numeric(k_folds)
    imputation_preds <- tibble()  # Empty tibble to collect preds for this imputation

      formula <- as.formula(paste("COVID ~", paste(predictors, collapse = " + ")))
      rf_model <- ranger(
        formula = formula,
        data = train_data,
        case.weights = train_data$SAMPWEIGHT,
        probability = TRUE
      )

      pred_probs <- predict(rf_model, data = test_data)$predictions[, 2]
      actual <- test_data$COVID

      # Store predictions
      imputation_preds <- bind_rows(imputation_preds, tibble(actual = actual, predicted = pred_probs))

      # AUC
      auc_result <- roc(actual, pred_probs)$auc
      fold_auc[fold] <- auc_result
    }

    auc_list[i] <- mean(fold_auc)
    predictions_list[[i]] <- imputation_preds
  }

  list(
    auc_per_imputation = auc_list,
    overall_auc = mean(auc_list),
    predictions = predictions_list  # List of tibbles with actual + predicted
  )
}
# Assuming you have your imputed_data and a vector of predictors
results_with_preds <- cross_validation_imputed_predictions(
  imputed_data = data_imputed,
  predictors = c("EXERCISE", "SLEEP", "DEPRESSED", "TROUBLE_SLEEPING", "SOCIAL")
)

# View results
results$auc_per_imputation  # AUC for each imputed dataset
results$overall_auc         # Average AUC across imputations

# Create calibration plot from predictions across imputations
plot_calibration_multiple <- function(predictions_list, n_bins = 10, title = "Calibration Plot") {
  calib_data_all <- map2_dfr(predictions_list, seq_along(predictions_list), function(df, i) {
    df %>%
      mutate(bin = ntile(predicted, n_bins)) %>%
      group_by(bin) %>%
      summarise(
        mean_pred = mean(predicted),
        obs_rate = mean(actual),
        .groups = "drop"
      ) %>%
      mutate(imputation = paste0("Imputation ", i))
  })

  ggplot(calib_data_all, aes(x = mean_pred, y = obs_rate, color = imputation)) +
    geom_line() +
    geom_point(size = 2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
    scale_x_continuous("Expected probability", limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    scale_y_continuous("Observed proportion", limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    labs(title = title, color = "Imputation") +
    theme_minimal()
}

# Plot it
plot_cal <- plot_calibration_multiple(results_with_preds$predictions, title = "Calibration Plot: RF across 5 Imputations")
print(plot_cal)

# Save it
ggsave(here("results", "figures", "rf_calibration_all_imputations.png"), plot_cal, width = 8, height = 6, dpi = 300)

@@

# Save the table as a PNG
here()
gtsave(model_table, here("results", "figures", "model_table.png"))

# Store your AUC lists here
aauc_multivar <- results$auc_per_imputation  # From multivariate logistic regression with CV
auc_lasso     <- unlist(auc_list)                      # From your LASSO section
auc_rf        <- auc_list                              # From your final RF model section

# Create a data frame with AUC values and model type
auc_df <- data.frame(
  Imputation = rep(1:5, 3),
  AUC = c(auc_multivar, auc_lasso, auc_rf),
  Model = rep(c("Multivariate Logistic Regression", "LASSO", "Random Forest"), each = 5)
)

# Plot: AUCs across imputations by model
aucs <- ggplot(auc_df, aes(x = Model, y = AUC, color = Model)) +
  geom_jitter(width = 0.2, size = 2) +         # dots for individual AUCs
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  stat_summary(fun = mean, geom = "text", aes(label = round(..y.., 3)), 
               vjust = -1, color = "black", fontface = "bold") +
  labs(title = "AUC Across Imputations by Model",
       x = "Model",
       y = "AUC") +
  theme_minimal()

# Save the table as a PNG
here()
ggsave(
  filename = here("results", "figures", "auc_plot.png"),
  plot = aucs,  # replace with your plot object name
  width = 8,
  height = 6,
  dpi = 300
)

Now, we will make expected versus observed plots.
```{r}
plot_calibration_multiple <- function(actual, predicted_list, n_bins = 10, title = "Calibration Plot") {
  # Combine calibration data across imputations
  calib_data_all <- map2_dfr(predicted_list, seq_along(predicted_list), function(predicted, i) {
    tibble(actual = actual, predicted = predicted) %>%
      mutate(bin = ntile(predicted, n_bins)) %>%
      group_by(bin) %>%
      summarise(
        mean_pred = mean(predicted),
        obs_rate = mean(actual),
        .groups = "drop"
      ) %>%
      mutate(imputation = paste0("Imputation ", i))
  })

  # Plot
  ggplot(calib_data_all, aes(x = mean_pred, y = obs_rate, color = imputation)) +
    geom_line() +
    geom_point(size = 2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
    scale_x_continuous("Expected probability", limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    scale_y_continuous("Observed proportion", limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    labs(title = title, color = "Imputation") +
    theme_minimal()
}

# Example usage:
plot_cal <- plot_calibration_multiple(
  actual = test_data$COVID,
  predicted_list = pred_probs_list,  # this should be a list of 5 vectors
  title = "Random Forest Calibration Across 5 Imputations"
)

# Display plot
print(plot_cal)

# Save it
ggsave(
  filename = here("results", "figures", "rf_calibration_all_imputations.png"),
  plot = plot_cal,
  width = 8,
  height = 6,
  dpi = 300
)

### 2025-05-01 products/manuscript/Manuscript.qmd
Part 6 updates

To explore the data further, a multivariate model, LASSO regression, and a third advanced model (replace this) were ran with cross validation.

>>>>

To better understand the relationship between COVID-19 test status and mental health factors, three new models were constructed: a multivariate logistic regression model, LASSO regression, and a random forest model. Each model used the imputed data and five fold cross validation to affect predictive performance. All models accounted for sample wighting, but not clustering. Notably, survey design could not be fully accounted for due to the complexity of the LASSO and random forest models, making results not interpretable to the country-wide level, rather, predictive results are relevant within the NHIS population, which accounts for 110,216 individuals.

All models included COVID-19 test status as the outcome and all mental helath variables as predictors (hours of sleep, touble sleeping, feeling depressed, meeting exercise guidelines, and amount of social support). The multivariate logistic regression was chosen for it's ease of interpretability and to serve as a baseline against the more advanced models. The LASSO model was chosen for it's ability to penalize predictors to indicate which mental health factors were most important. The random forest model was used to capture potential complex relationships between variables.

To evalaute the performance of these models, the area under the curve (AUC) was calculated. This metric was chosen due to it's discrimination between binary outcome of the COVID-19 variable. Higher AUC indicates better model preformance.

```{r}
#| label: fig-aucplot
#| fig-cap: "AUC comparison plot"
#| echo: FALSE
here()
knitr::include_graphics(here("results","figures", "auc_plot.png"))
```
